Responsible Machine Learning Techniques\\\large{Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2019-12-20 article-information-2019.tex
finish first draft of 2.1

Ridge functions are reminiscent of basis functions ... distinctive simplistic functions

@@

In combination, Figure \ref{fig:sim_xnn_glob_ridge} \textbf{A}, \textbf{B}, and \textbf{C} help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$.

Deep SHAP values are reported in the probability space, after the application of the logit link function. They are also calculated from the projection layer of $g^\text{XNN}$.

@@

@@

11.1\% fewer control group 1 members receive the favorable offer under the ME column in Table \ref{tab:sim_disc}.  Of note is that 11.1\% is not a meaningful difference without context.  If the population of control group 1 and control group 2 were substantially similar in relevant characteristics, 11.1\% could represent an extremely large difference and would require remediation.  But if they represent substantially different populations, then 11.1\% could represent a reasonable deviation from parity.  As an example, if a lending institution that has traditionally focused on high credit quality clients were to expand into previously under-banked communities, an 11.1\% class-control difference in loan approval rates might be expected because the average credit quality of the new population would be lower than that of the existing population. Protected group 1’s AIR under $g^\text{XNN}$ is 0.737, is below the EEOC 4/5ths rule.  It is also highly statistically significant (not shown, but available in resources discussed in Section \ref{sec:soft}), which together would indicate that there may be evidence of illegal DI.  As with ME and other measures, the reasonableness of this disparity is not clear outside of context.  However, most regulated institutions that do perform discrimination analyses would find an AIR of this magnitude concerning and warranting further review.

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g. numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced AUC values of $\sim0.81$ (not shown, but available in resources discussed in Section \ref{sec:soft}).

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower,  exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Section \ref{sec:soft}). However, fluctuations in ICE can also be caused by overfitting or leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

>>>>

Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge} \textbf{A} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge} \textbf{B} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{A} and \ref{fig:sim_xnn_glob_ridge} \textbf{B} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

@@

\noindent $n_k$ subnetworks are plotted across the output values of their associated $\sum_j\beta_{k,j}x_j$ projection layer hidden units, and color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{B} and \ref{fig:sim_xnn_glob_ridge} \textbf{C} link the $\beta_j$ weights to their $n_k$ subnetworks. Most of the heavily utilized $n_k$ subnetworks have sparse weights in their $\sum_j\beta_{k,j}x_j$ projection layer hidden units. In particular, subnetwork $n_1$ appears to be almost solely a function of $X_{\text{Friedman}, 3}$ and appears to exhibit the expected quadratic behavior for $X_{\text{Friedman}, 3}$. Subnetworks $n_7$, $n_8$, and $n_9$ appear to be most associated with the globally important $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ features and do display some noticeable sinusoidal characteristics. Conversely, subnetwork $n_4$ seems to be a linear combination of all the original input $X_j$ features. As a whole, Figure \ref{fig:sim_xnn_glob_ridge} \textbf{A}, \textbf{B}, and \textbf{C} help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$.

Local Deep SHAP feature importance in Figure \ref{fig:sim_xnn_loc} supplements the global interpretability of $g^\text{XNN}$ displayed in Figures \ref{fig:sim_xnn_glob} and \ref{fig:sim_xnn_glob_ridge}. Local Deep SHAP values are extracted from the projection layer of $g^\text{XNN}$ and reported in the probability space. Deep SHAP values can be calculated for any arbitrary $g^\text{XNN}(\mathbf{x})$, allowing for detailed, local summarization of individual model predictions.

@@

\noindent As expected, Deep SHAP values generally increase from the 10\textsuperscript{th} percentile of $g^\text{XNN}(\mathbf{X})$ to the 90\textsuperscript{th} percentile of $g^\text{XNN}(\mathbf{X})$, with primarily important global drivers of model behavior contributing to the selected local $g^\text{XNN}(\mathbf{x}^{(i)})$ predictions.

@@

11.1\% fewer control group 1 members receive the favorable offer under the ME column in Table \ref{tab:sim_disc}.  Of note is that 11.1\% is not a meaningful difference without context.  If the population of control group 1 and control group 2 were substantially similar in relevant characteristics, 11.1\% could represent an extremely large difference and would require remediation.  But if they represent substantially different populations, then 11.1\% could represent a reasonable deviation from parity.  As an example, if a lending institution that has traditionally focused on high credit quality clients were to expand into previously under-banked communities, an 11.1\% class-control difference in loan approval rates might be expected because the average credit quality of the new population would be lower than that of the existing population. Protected group 1’s AIR under $g^\text{XNN}$ is 0.737, is below the EEOC 4/5ths rule.  It is also highly statistically significant (not shown, but available in resources discussed in Subsection \ref{ssec:soft}), which together would indicate that there may be evidence of illegal DI.  As with ME and other measures, the reasonableness of this disparity is not clear outside of context.  However, most regulated institutions that do perform discrimination analyses would find an AIR of this magnitude concerning and warranting further review.

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g. numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced AUC values of $\sim0.81$ (not shown, but available in resources discussed in Subsection \ref{ssec:soft}).

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower,  exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Subsection \ref{ssec:soft}). However, fluctuations in ICE can also be caused by overfitting or leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.
